Synthesizing Symmetric Lenses

### 2018-07-11 papers/SymmetricLenses/main.tex
fixed capitalization error in boomerang citation that was generaing bibtex errors
removed information extraction from benchmark description

We implemented the \GreedySynth and \Expand algorithms and evaluated their
effectiveness at synthesizing lenses on a 
benchmark suite of 45 problems, including data cleaning tasks taken
from Flash Fill~\cite{gulwani-popl-2014}, information 
extraction tasks \ksf{Dave wants to change information extraction to
soemthing else.} taken from Augeas~\cite{Augeas}, and ad hoc format synchronization tasks
taken from a variety of other locations\footnote{The majority of these format
  synchronization tasks were found from the following index:
  \url{http://fileformats.archiveteam.org/wiki/Electronic_File_Formats}}.

In summary, our main contributions are:
\begin{itemize}
\item We define simple symmetric lenses, which have
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We give a language for defining such lenses and describe how to
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds
  (\S\ref{sec:relationship}).
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on regular
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend Boomerang with simple symmetric lenses,
  showing that such lenses can integrate smoothly with 
  other lens extensions such as quotient lenses and matching
  lenses.  We evaluate our
  implementation on a set of 45 lens synthesis benchmarks drawn from
  data cleaning, \ksf{Dave wants to change information extraction to
soemthing else.} information 
  extraction, and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We finish the paper with a discussion of related work (\S\ref{sec:related}) and
our concluding thoughts (\S\ref{sec:conc}).

@@

But what is a ``likely'' satisfying lens? We propose the following
heuristic: a satisfying
lens is more likely if it connects more data between the two
formats, \EG, by using an identity rather than a disconnect lens.  The
likeliest satisfying lenses would connect as much data as
possible. More formally, we define the likelihood of a
satisfying lens as the expected number of bits required to recover
data in one format from synchronized data in the other.  Higher
likelihoods correspond to lower numbers of bits. In more
detail, we say that two strings $s$ and $t$ are \emph{synchronized
according to lens $l$} if $l.\PutR\ s\ t = t$ and $l.\PutL\ t\ s = s$.\footnote{
The notation $l.\PutR$ extracts the \PutR function from simple
symmetric lens $l$; similarly for \PutL.}  We say that we
can \emph{recover} $s$ from $t$ using bits $b$ and lens $l$ if we can
reconstruct $s$ from $t$, $b$, and $l$.  For example, given
the \IdentityLens{} lens, we can recover $s$ from $t$ using no bits
because, in this case, $s$ is just $t$.  In contrast, given
the \Disconnect{} lens, we need enough bits to fully encode $s$ in
order to recover it from $t$ because $t$ contains no useful
information.  This calculation shows that if both \IdentityLens{}
and \Disconnect{} are satisfying lenses, then \IdentityLens{} is more
likely.

@@

\ksf{Consider integrating more detailed discussion of SREs that was
formerly in the intro.  Commented out here for now.}
%%To implement this intuition, we turn to \emph{stochastic regular
%%expressions} (SREs). Stochastic 
%%regular expressions define a set of strings, as do ordinary regular expressions,
%%but they also provide the expected distribution of strings within that set. Such
%%distributions allow us to compute the amount of information gained when
%%presented an element of the set. For example, the language of the SRE
%%$\lstinline{"a"} |_{0.5}
%%\lstinline{"b"}$ contains two elements (the strings \lstinline{"a"} and \lstinline{"b"}).
%%The probability 0.5 annotating the union indicates that both elements are
%%equally likely to appear. Consequently, if we observe an element of that
%%language (either \lstinline{"a"} or
%%\lstinline{"b"}), we will gain one bit of information. Said another way, one bit, together with
%%the SRE, suffices to recover the string in question. Interestingly, if the
%%probability annotation were anything but 0.5, the expected information content
%%(or entropy) would actually be less than 1 (given a stream of such choices, on
%%average we could encode the the contents of stream in fewer bits than the the
%%number of choices present). On the other hand, consider the language of the SRE
%%\lstinline{c}. It contains just one element (the string $c$). If we observe that
%%our algorithm generates $\lstinline{"c"}$, we gain zero bits of information ---
%%the string was already determined. Using this foundation, a lens $\Lens$ that
%%synchronizes data between $\Regex$ and $\RegexAlt$ is ``likely,''
%%in a technical sense, if given a string $s$ in the source $S$, very little information
%%is needed to determine which synchronized string $t$ is present in the target
%%$T$.

This infrastructure gives us what we need to find likely satisfying lenses from
a collection of satisfying lenses. All such lenses must synchronize data between
the same two SREs $\Regex$ and $\RegexAlt$. For any such lens $\Lens$, we can
calculate the expected number of bits required to recover a string $t$ in
$\LanguageOf{\RegexAlt}$ from a synchronized string $s$ in
$\LanguageOf{\Regex}$. This expectation is the \emph{conditional entropy of
  $\Regex$ given $\RegexAlt$ and $\Lens$}, formally $\sum_{s \in
  \Regex}\ProbabilityOf{\Regex}{s}*\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher cost to lenses where knowing the string present in \Regex
provides little information about what string is present in \RegexAlt. With this
metric, bijections have zero cost---given a string $s \in \Regex$, we know
exactly what string is present in $\RegexAlt$, as exactly one string is
synchronized to it.

To obtain SREs from the plain regular expressions that we expect users to
supply, we use a heurstic that attempts to assign probability annotations that
give each string in the language equal probablity. Our algorithm could also take
as input a stochastic regular expression rather than a plain one, which allows
the user to specify a data-specific probability distribution, either manually
written or inferred from a large dataset. We found our heuristically chosen
distribution works well in practice.

@@

We now provide the typing derivations and semantics for these lenses. We do not
include the semantics for combinators equivalent to those presented in prior
work~\cite{symmetric-lenses,boomerang}. Furthermore, we only present \CreateR
and \PutR when \CreateL and \PutL are symmetric. Full details are present in the
appendix. For simplicity of presentation, we define the typing derivations and
semantics only over unambiguous regular expressions, even though some lenses can
be defined over ambiguous regular expressions (for example, the identity lens
doesn't need any ambiguity constraints). We believe relaxing this restriction is
unproblematic.

>>>>

We implemented the \GreedySynth and \Expand algorithms and evaluated their
effectiveness at synthesizing lenses on a 
benchmark suite of 45 problems, including data cleaning tasks taken
from Flash Fill~\cite{gulwani-popl-2014}, view maintenance
tasks taken from Augeas~\cite{Augeas}, and ad hoc format synchronization tasks
taken from a variety of other locations\footnote{The majority of these format
  synchronization tasks were found from the following index:
  \url{http://fileformats.archiveteam.org/wiki/Electronic_File_Formats}}.

In summary, our main contributions are:
\begin{itemize}
\item We define simple symmetric lenses, which have
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We give a language for defining such lenses and describe how to
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds
  (\S\ref{sec:relationship}).
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on regular
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend Boomerang with simple symmetric lenses,
  showing that such lenses can integrate smoothly with 
  other lens extensions such as quotient lenses and matching
  lenses.  We evaluate our
  implementation on a set of 45 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We finish the paper with a discussion of related work (\S\ref{sec:related}) and
our concluding thoughts (\S\ref{sec:conc}).

@@

But what is a ``likely'' satisfying lens? We propose the following
heuristic: A satisfying
lens is more likely if it connects more data between the two
formats, \EG, by using an identity rather than a disconnect lens.  The
likeliest satisfying lenses would connect as much data as
possible. More formally, we define the likelihood of a
satisfying lens as the expected number of bits required to recover
data in one format from synchronized data in the other.  Higher
likelihoods correspond to lower numbers of bits. In more
detail, we say that two strings $s$ and $t$ are \emph{synchronized
according to lens $l$} if $l.\PutR\ s\ t = t$ and $l.\PutL\ t\ s = s$.\footnote{
The notation $l.\PutR$ extracts the \PutR function from simple
symmetric lens $l$; similarly for \PutL.}  We say that we
can \emph{recover} $s$ from $t$ using bits $b$ and lens $l$ if we can
reconstruct $s$ from $t$, $b$, and $l$.  For example, given
the \IdentityLens{} lens, we can recover $s$ from $t$ using no bits
because, in this case, $s$ is just $t$.  In contrast, given
the \Disconnect{} lens, we need enough bits to fully encode $s$ in
order to recover it from $t$ because $t$ contains no useful
information.  This calculation shows that if both \IdentityLens{}
and \Disconnect{} are satisfying lenses, then \IdentityLens{} is more
likely.

@@

%% \ksf{Consider integrating more detailed discussion of SREs that was
%%formerly in the intro.  Commented out here for now.}
%%To implement this intuition, we turn to \emph{stochastic regular
%%expressions} (SREs). Stochastic 
%%regular expressions define a set of strings, as do ordinary regular expressions,
%%but they also provide the expected distribution of strings within that set. Such
%%distributions allow us to compute the amount of information gained when
%%presented an element of the set. For example, the language of the SRE
%%$\lstinline{"a"} |_{0.5}
%%\lstinline{"b"}$ contains two elements (the strings \lstinline{"a"} and \lstinline{"b"}).
%%The probability 0.5 annotating the union indicates that both elements are
%%equally likely to appear. Consequently, if we observe an element of that
%%language (either \lstinline{"a"} or
%%\lstinline{"b"}), we will gain one bit of information. Said another way, one bit, together with
%%the SRE, suffices to recover the string in question. Interestingly, if the
%%probability annotation were anything but 0.5, the expected information content
%%(or entropy) would actually be less than 1 (given a stream of such choices, on
%%average we could encode the the contents of stream in fewer bits than the the
%%number of choices present). On the other hand, consider the language of the SRE
%%\lstinline{c}. It contains just one element (the string $c$). If we observe that
%%our algorithm generates $\lstinline{"c"}$, we gain zero bits of information ---
%%the string was already determined. Using this foundation, a lens $\Lens$ that
%%synchronizes data between $\Regex$ and $\RegexAlt$ is ``likely,''
%%in a technical sense, if given a string $s$ in the source $S$, very little information
%%is needed to determine which synchronized string $t$ is present in the target
%%$T$.

This infrastructure gives us what we need to find likely satisfying lenses from
a collection of satisfying lenses. All such lenses must synchronize data between
the same two SREs $\Regex$ and $\RegexAlt$. For any such lens $\Lens$, we can
calculate the expected number of bits required to recover a string $t$ in
$\LanguageOf{\RegexAlt}$ from a synchronized string $s$ in
$\LanguageOf{\Regex}$. This expectation is the \emph{conditional entropy of
  $\Regex$ given $\RegexAlt$ and $\Lens$}, formally $\sum_{s \in
  \Regex}\ProbabilityOf{\Regex}{s}*\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher entropy (or cost) to lenses where knowing the string
on one side 
provides little information about the string on the other side. 
It assigns zero entropy to bijections because given a string $s \in \Regex$,
the bijection exactly determines
the corresponding string in 
$\RegexAlt$.

To obtain SREs from the plain regular expressions that we expect users to
supply, we use a heuristic that attempts to assign probability annotations that
give each string in the language equal probablity. Our algorithm could also take
as input a stochastic regular expression rather than a plain one,
which would allow
the user to specify a data-specific probability distribution, either manually
calculated or inferred from a large dataset. We found our heuristic
works well in practice, so we have not explored these options.

@@

We now provide the typing derivations and semantics for these lenses. We do not
include the semantics for combinators equivalent to those presented in prior
work~\cite{symmetric-lenses,Boomerang}. Furthermore, we only present \CreateR
and \PutR when \CreateL and \PutL are symmetric. Full details are present in the
appendix. For simplicity of presentation, we define the typing derivations and
semantics only over unambiguous regular expressions, even though some lenses can
be defined over ambiguous regular expressions (for example, the identity lens
doesn't need any ambiguity constraints). We believe relaxing this restriction is
unproblematic.
